
---
output: pdf_document
---


```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```
	  
# Multiple imputation for missing data  

## MSDS 7333 - Quantifying the World - Unit 06 Case Study  
### Patrick McDevitt  
#### 10-Jun-2018  

***  
	  
##Abstract  

##Introduction  

##Literature review  

##Methods  


```{r initializations, include = FALSE}

rm(list = ls())

setwd("~/_ds/_smu/_src/qtw/unit_06_geo_loc")

options(digits = 2)


```

```{r utility functions, include = FALSE}

processLine = function(x)
{
	tokens = strsplit(x, "[;=,]")[[1]]
	
	if (length(tokens) == 10) return(NULL)
 
	tmp = matrix(tokens[ - (1:10) ],
	             ncol = 4,
	             byrow = TRUE)
	
	cbind(matrix(tokens[c(2, 4, 6:8, 10)],
	             nrow(tmp),
	             6, 
	             byrow = TRUE),
	      tmp)
}

roundOrientation = function(angles)
{
	refs = seq(0, by = 45, length  = 9)
	q = sapply(angles, function(o) which.min(abs(o - refs)))
	c(refs[1:8], 0)[q]
}



readData = function(
            filename = 'offline.final.trace.txt', 
	        subMacs = c("00:0f:a3:39:e1:c0", "00:0f:a3:39:dd:cd", "00:14:bf:b1:97:8a",
	                     "00:14:bf:3b:c7:c6", "00:14:bf:b1:97:90", "00:14:bf:b1:97:8d",
	                     "00:14:bf:b1:97:81"))
{
	  txt = readLines(filename)
	  lines = txt[ substr(txt, 1, 1) != "#" ]
	  tmp = lapply(lines, processLine)
	  offline = as.data.frame(do.call("rbind", tmp), 
	                          stringsAsFactors= FALSE) 
	  
	  names(offline) = c("time", "scanMac", 
	                     "posX", "posY", "posZ", "orientation", 
	                     "mac", "signal", "channel", "type")
	  
	   # keep only signals from access points
	  offline = offline[ offline$type == "3", ]
	  
	  # drop scanMac, posZ, channel, and type - no info in them
	  dropVars = c("scanMac", "posZ", "channel", "type")
	  offline = offline[ , !( names(offline) %in% dropVars ) ]
	  
	  # drop more unwanted access points
	  offline = offline[ offline$mac %in% subMacs, ]
	  
	  # convert numeric values
	  numVars = c("time", "posX", "posY", "orientation", "signal")
	  offline[ numVars ] = lapply(offline[ numVars ], as.numeric)

	  # convert time to POSIX
	  offline$rawTime = offline$time
	  offline$time = offline$time/1000
	  class(offline$time) = c("POSIXt", "POSIXct")
	  
	  # round orientations to nearest 45
	  offline$angle = roundOrientation(offline$orientation)
	    
	  return(offline)
}

```


# ...   -=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-
# ...   code from : http://rdatasciencecases.org/GeoLoc/code.R
# ...   -=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-

```{r read data, include = FALSE}


# ...   read in raw data file

txt = readLines("offline.final.trace.txt")

# ...   count number comment lines

sum(substr(txt, 1, 1) == "#")

# ...   confirm length of text
length(txt)

# ...   drop comment lines

lines = txt[ substr(txt, 1, 1) != "#" ]

# ...   split on data fields, format into untidy data frame

#options(error = recover, warn = 1)

tmp = lapply(lines, processLine)
offline = as.data.frame(do.call("rbind", tmp), 
	                      stringsAsFactors = FALSE)

dim(offline)

```



```{r clean1, include = FALSE}

# ...   add some column names

names(offline) = c("time", "scanMac", "posX", "posY", "posZ", 
	                 "orientation", "mac", "signal", 
	                 "channel", "type")

# ...   convert to numeric

numVars = c("time", "posX", "posY", "posZ", 
	          "orientation", "signal")
offline[ numVars ] =  lapply(offline[ numVars ], as.numeric)

# ... remove ad hoc measurements

offline = offline[ offline$type == "3", ]

# ... remove type variable from data frame

offline = offline[ , "type" != names(offline) ]
dim(offline)

# ... scale time to seconds and set as time class

offline$rawTime = offline$time
offline$time = offline$time/1000
class(offline$time) = c("POSIXt", "POSIXct")

# ... verify data types are correct

unlist(lapply(offline, class))

# ... simple verification of data correct

summary(offline[, numVars])

# ... convert character variables to factors

summary(sapply(offline[ , c("mac", "channel", "scanMac")],
	              as.factor))

# ... eliminate a few rare observations

offline = offline[ , !(names(offline) %in% c("scanMac", "posZ"))]

# ... check length of orientation vector

length(unique(offline$orientation))

```



```{r clean2, include = FALSE}

# ...   round orientation to nearest 45 deg 

plot(ecdf(offline$orientation))

pdf(file = "Geo_ECDFOrientation.pdf", width = 10, height = 7)
oldPar = par(mar = c(4, 4, 1, 1))
plot(ecdf(offline$orientation), pch = 19, cex = 0.3,
	   xlim = c(-5, 365), axes = FALSE,
	   xlab = "orientation", ylab = "Empirical CDF", main = "")
box()
axis(2)
axis(side = 1, at = seq(0, 360, by = 45))
par(oldPar)
dev.off()

pdf(file = "Geo_DensityOrientation.pdf", width = 10, height = 5)
oldPar = par(mar = c(4, 4, 1, 1))
plot(density(offline$orientation, bw = 2), 
 xlab = "orientation", main = "")
par(oldPar)
dev.off()

offline$angle = roundOrientation(offline$orientation)

with(offline, boxplot(orientation ~ angle,
xlab = "nearest 45 degree angle",
ylab="orientation"))

pdf(file = "Geo_BoxplotAngle.pdf", width = 10)
oldPar = par(mar = c(4, 4, 1, 1))
plot(ecdf(offline$orientation), pch = 19, cex = 0.3,
	   xlim = c(-5, 365), axes = FALSE,
	   xlab = "orientation", ylab = "Empirical CDF", main = "")
box()
axis(2)
axis(side = 1, at = seq(0, 360, by = 45))

par(oldPar)
dev.off()

```


```{r mac address, include = FALSE}

c(length(unique(offline$mac)), length(unique(offline$channel)))

# ... check the counts of observations for the various MAC addresses

table(offline$mac)

# ...   We look up these MAC addresses at the http://coffer.com/
# ...   mac_find/ site to find the vendor addresses that begin with 00:14:bf belong to Linksys
# ...   devices, those beginning with 00:0f:a3 belong to Alpha Networks, and Lancom devices
# ...   start with 00:a0:57 (see Figure 1.4). We do have 5 devices with an address that begins
# ...   00:14:bf, which matches with the Linksys count from the documentation. However, none
# ...   of our MAC addresses begin with 00:a0:57 so there is a discrepancy with the documentation


# ... keep the records from the top 7 device counts

subMacs = names(sort(table(offline$mac), decreasing = TRUE))[1:7]
offline = offline[ offline$mac %in% subMacs, ]

# ...   confirm there is at least 1 non-zero entry per row

macChannel = with(offline, table(mac, channel))
apply(macChannel, 1, function(x) sum(x > 0))

# ...   eliminate channel from dataframe

offline = offline[ , "channel" != names(offline)]

```


```{r hand held device, include = FALSE}

# ...   create list for each location

locDF = with(offline, 
	           by(offline, list(posX, posY), function(x) x))
length(locDF)

# ...   check for empty elements

sum(sapply(locDF, is.null))

# ...   drop unneeded null elements

locDF = locDF[ !sapply(locDF, is.null) ]

length(locDF)

# ...   check number observations at each location

locCounts = sapply(locDF, nrow)
locCounts = sapply(locDF, 
	                 function(df) 
	                   c(df[1, c("posX", "posY")], count = nrow(df)))

# ...   confirm matrix with 3 rows

class(locCounts)
dim(locCounts)

# ... examine a few locations

locCounts[ , 1:8]

# ...   make a plot of number of observations / location

#pdf(file = "Geo_XYByCount.pdf", width = 10)
#oldPar = par(mar = c(3.1, 3.1, 1, 1))

locCounts = t(locCounts)

plot(locCounts, type = "n", xlab = "", ylab = "")
text(locCounts, labels = locCounts[,3], cex = .8, srt = 45)

#par(oldPar)
dev.off()

```


```{r master read, include = FALSE}

offlineRedo = readData()

identical(offline, offlineRedo)

```


```{r signal strength, include = FALSE}

pdf(file = "Geo_BoxplotSignalByMacAngle.pdf", width = 7)
oldPar = par(mar = c(3.1, 3, 1, 1))

library(lattice)
bwplot(signal ~ factor(angle) | mac, data = offline, 
	     subset = posX == 2 & posY == 12 
	              & mac != "00:0f:a3:39:dd:cd", 
	     layout = c(2,3))

par(oldPar)
dev.off()

summary(offline$signal)

pdf(file = "Geo_DensitySignalByMacAngle.pdf", width = 8, height = 12)
oldPar = par(mar = c(3.1, 3, 1, 1))

densityplot( ~ signal | mac + factor(angle), data = offline,
	           subset = posX == 24 & posY == 4 & 
	                       mac != "00:0f:a3:39:dd:cd",
	           bw = 0.5, plot.points = FALSE)

par(oldPar)
dev.off()

#offline = offline[ offline$mac != "00:0f:a3:39:dd:cd", ]

offline$posXY = paste(offline$posX, offline$posY, sep = "-")

byLocAngleAP = with(offline, 
	                  by(offline, list(posXY, angle, mac), 
	                     function(x) x))

signalSummary = 
	lapply(byLocAngleAP,            
	       function(oneLoc) {
	         ans = oneLoc[1, ]
	         ans$medSignal = median(oneLoc$signal)
	         ans$avgSignal = mean(oneLoc$signal)
	         ans$num = length(oneLoc$signal)
	         ans$sdSignal = sd(oneLoc$signal)
	         ans$iqrSignal = IQR(oneLoc$signal)
	         ans
	         })

offlineSummary = do.call("rbind", signalSummary)     

pdf(file = "Geo_BoxplotSignalSDByAvg.pdf", width = 10)
oldPar = par(mar = c(3.1, 3, 1, 1))

breaks = seq(-90, -30, by = 5)
bwplot(sdSignal ~ cut(avgSignal, breaks = breaks),
	     data = offlineSummary, 
	     subset = mac != "00:0f:a3:39:dd:cd",
	     xlab = "Mean Signal", ylab = "SD Signal")

par(oldPar)
dev.off()

pdf(file = "Geo_ScatterMean-Median.pdf", width = 10)
oldPar = par(mar = c(4.1, 4.1, 1, 1))

with(offlineSummary,
	   smoothScatter((avgSignal - medSignal) ~ num,
	                 xlab = "Number of Observations", 
	                 ylab = "mean - median"))
abline(h = 0, col = "#984ea3", lwd = 2)

lo.obj = 
	with(offlineSummary,
	     loess(diff ~ num, 
	           data = data.frame(diff = (avgSignal - medSignal),
	                             num = num)))

lo.obj.pr = predict(lo.obj, newdata = data.frame(num = (70:120)))
lines(x = 70:120, y = lo.obj.pr, col = "#4daf4a", lwd = 2)

par(oldPar)
dev.off()

```

 
oneAPAngle = subset(offlineSummary, 
	                  mac == subMacs[5] & angle == 0)


library(fields)
smoothSS = Tps(oneAPAngle[, c("posX","posY")], 
	             oneAPAngle$avgSignal)

vizSmooth = predictSurface(smoothSS)

plot.surface(vizSmooth, type = "C")

points(oneAPAngle$posX, oneAPAngle$posY, pch=19, cex = 0.5)

surfaceSS = function(data, mac, angle = 45) {
	require(fields)
	oneAPAngle = data[ data$mac == mac & data$angle == angle, ]
	smoothSS = Tps(oneAPAngle[, c("posX","posY")], 
	               oneAPAngle$avgSignal)
	vizSmooth = predictSurface(smoothSS)
	plot.surface(vizSmooth, type = "C", 
	             xlab = "", ylab = "", xaxt = "n", yaxt = "n")
	points(oneAPAngle$posX, oneAPAngle$posY, pch=19, cex = 0.5) 
}

parCur = par(mfrow = c(2,2), mar = rep(1, 4))

mapply(surfaceSS, mac = subMacs[ rep(c(5, 1), each = 2) ], 
	     angle = rep(c(0, 135), 2),
	     data = list(data = offlineSummary))
 
par(parCur)

offlineSummary = subset(offlineSummary, mac != subMacs[2])

AP = matrix( c( 7.5, 6.3, 2.5, -.8, 12.8, -2.8,  
	              1, 14, 33.5, 9.3,  33.5, 2.8),
	          ncol = 2, byrow = TRUE,
	          dimnames = list(subMacs[ -2 ], c("x", "y") ))

AP

diffs = offlineSummary[ , c("posX", "posY")] - 
	        AP[ offlineSummary$mac, ]

offlineSummary$dist = sqrt(diffs[ , 1]^2 + diffs[ , 2]^2)

xyplot(signal ~ dist | factor(mac) + factor(angle), 
	     data = offlineSummary, pch = 19, cex = 0.3,
	     xlab ="distance")

pdf(file="Geo_ScatterSignalDist.pdf", width = 7, height = 10)
oldPar = par(mar = c(3.1, 3.1, 1, 1))
library(lattice)
xyplot(signal ~ dist | factor(mac) + factor(angle), 
	     data = offlineSummary, pch = 19, cex = 0.3,
	     xlab ="distance")
par(oldPar)
dev.off()

macs = unique(offlineSummary$mac)
online = readData("Data/online.final.trace.txt", subMacs = macs)

online$posXY = paste(online$posX, online$posY, sep = "-")

length(unique(online$posXY))

tabonlineXYA = table(online$posXY, online$angle)
tabonlineXYA[1:6, ]

keepVars = c("posXY", "posX","posY", "orientation", "angle")
byLoc = with(online, 
	           by(online, list(posXY), 
	              function(x) {
	                ans = x[1, keepVars]
	                avgSS = tapply(x$signal, x$mac, mean)
	                y = matrix(avgSS, nrow = 1, ncol = 6,
	                      dimnames = list(ans$posXY, names(avgSS)))
	                cbind(ans, y)
	              }))

onlineSummary = do.call("rbind", byLoc)  

dim(onlineSummary)

names(onlineSummary)
m = 3; angleNewObs = 230
refs = seq(0, by = 45, length  = 8)
nearestAngle = roundOrientation(angleNewObs)
	
if (m %% 2 == 1) {
	angles = seq(-45 * (m - 1) /2, 45 * (m - 1) /2, length = m)
} else {
	m = m + 1
	angles = seq(-45 * (m - 1) /2, 45 * (m - 1) /2, length = m)
	if (sign(angleNewObs - nearestAngle) > -1) 
	  angles = angles[ -1 ]
	else 
	  angles = angles[ -m ]
}
angles = angles + nearestAngle
angles[angles < 0] = angles[ angles < 0 ] + 360
angles[angles > 360] = angles[ angles > 360 ] - 360

offlineSubset = 
	offlineSummary[ offlineSummary$angle %in% angles, ]

reshapeSS = function(data, varSignal = "signal", 
	                   keepVars = c("posXY", "posX","posY")) {
	byLocation =
	  with(data, by(data, list(posXY), 
	                function(x) {
	                  ans = x[1, keepVars]
	                  avgSS = tapply(x[ , varSignal ], x$mac, mean)
	                  y = matrix(avgSS, nrow = 1, ncol = 6,
	                             dimnames = list(ans$posXY,
	                                             names(avgSS)))
	                  cbind(ans, y)
	                }))

	newDataSS = do.call("rbind", byLocation)
	return(newDataSS)
}

trainSS = reshapeSS(offlineSubset, varSignal = "avgSignal")

selectTrain = function(angleNewObs, signals = NULL, m = 1){
	# m is the number of angles to keep between 1 and 5
	refs = seq(0, by = 45, length  = 8)
	nearestAngle = roundOrientation(angleNewObs)
	
	if (m %% 2 == 1) 
	  angles = seq(-45 * (m - 1) /2, 45 * (m - 1) /2, length = m)
	else {
	  m = m + 1
	  angles = seq(-45 * (m - 1) /2, 45 * (m - 1) /2, length = m)
	  if (sign(angleNewObs - nearestAngle) > -1) 
	    angles = angles[ -1 ]
	  else 
	    angles = angles[ -m ]
	}
	angles = angles + nearestAngle
	angles[angles < 0] = angles[ angles < 0 ] + 360
	angles[angles > 360] = angles[ angles > 360 ] - 360
	angles = sort(angles) 
	
	offlineSubset = signals[ signals$angle %in% angles, ]
	reshapeSS(offlineSubset, varSignal = "avgSignal")
}

train130 = selectTrain(130, offlineSummary, m = 3)

head(train130)

length(train130[[1]])

findNN = function(newSignal, trainSubset) {
	diffs = apply(trainSubset[ , 4:9], 1, 
	              function(x) x - newSignal)
	dists = apply(diffs, 2, function(x) sqrt(sum(x^2)) )
	closest = order(dists)
	return(trainSubset[closest, 1:3 ])
}

predXY = function(newSignals, newAngles, trainData, 
	                numAngles = 1, k = 3){
	
	closeXY = list(length = nrow(newSignals))
	
	for (i in 1:nrow(newSignals)) {
	  trainSS = selectTrain(newAngles[i], trainData, m = numAngles)
	  closeXY[[i]] = 
	    findNN(newSignal = as.numeric(newSignals[i, ]), trainSS)
	}

	estXY = lapply(closeXY, 
	               function(x) sapply(x[ , 2:3], 
	                                  function(x) mean(x[1:k])))
	estXY = do.call("rbind", estXY)
	return(estXY)
}
estXYk3 = predXY(newSignals = onlineSummary[ , 6:11], 
	               newAngles = onlineSummary[ , 4], 
	               offlineSummary, numAngles = 3, k = 3)

estXYk1 = predXY(newSignals = onlineSummary[ , 6:11], 
	               newAngles = onlineSummary[ , 4], 
	               offlineSummary, numAngles = 3, k = 1)

floorErrorMap = function(estXY, actualXY, trainPoints = NULL, AP = NULL){
	
	  plot(0, 0, xlim = c(0, 35), ylim = c(-3, 15), type = "n",
	       xlab = "", ylab = "", axes = FALSE)
	  box()
	  if ( !is.null(AP) ) points(AP, pch = 15)
	  if ( !is.null(trainPoints) )
	    points(trainPoints, pch = 19, col="grey", cex = 0.6)
	  
	  points(x = actualXY[, 1], y = actualXY[, 2], 
	         pch = 19, cex = 0.8 )
	  points(x = estXY[, 1], y = estXY[, 2], 
	         pch = 8, cex = 0.8 )
	  segments(x0 = estXY[, 1], y0 = estXY[, 2],
	           x1 = actualXY[, 1], y1 = actualXY[ , 2],
	           lwd = 2, col = "red")
}

trainPoints = offlineSummary[ offlineSummary$angle == 0 & 
	                            offlineSummary$mac == "00:0f:a3:39:e1:c0" ,
	                      c("posX", "posY")]

pdf(file="GEO_FloorPlanK3Errors.pdf", width = 10, height = 7)
oldPar = par(mar = c(1, 1, 1, 1))
floorErrorMap(estXYk3, onlineSummary[ , c("posX","posY")], 
	            trainPoints = trainPoints, AP = AP)
par(oldPar)
dev.off()

pdf(file="GEO_FloorPlanK1Errors.pdf", width = 10, height = 7)
oldPar = par(mar = c(1, 1, 1, 1))
floorErrorMap(estXYk1, onlineSummary[ , c("posX","posY")], 
	            trainPoints = trainPoints, AP = AP)
par(oldPar)
dev.off()

calcError = 
function(estXY, actualXY) 
	 sum( rowSums( (estXY - actualXY)^2) )

actualXY = onlineSummary[ , c("posX", "posY")]
sapply(list(estXYk1, estXYk3), calcError, actualXY)


v = 11
permuteLocs = sample(unique(offlineSummary$posXY))
permuteLocs = matrix(permuteLocs, ncol = v, 
	                   nrow = floor(length(permuteLocs)/v))

onlineFold = subset(offlineSummary, posXY %in% permuteLocs[ , 1])

reshapeSS = function(data, varSignal = "signal", 
	                   keepVars = c("posXY", "posX","posY"),
	                   sampleAngle = FALSE, 
	                   refs = seq(0, 315, by = 45)) {
	byLocation =
	  with(data, by(data, list(posXY), 
	                function(x) {
	                  if (sampleAngle) {
	                    x = x[x$angle == sample(refs, size = 1), ]}
	                  ans = x[1, keepVars]
	                  avgSS = tapply(x[ , varSignal ], x$mac, mean)
	                  y = matrix(avgSS, nrow = 1, ncol = 6,
	                             dimnames = list(ans$posXY,
	                                             names(avgSS)))
	                  cbind(ans, y)
	                }))

	newDataSS = do.call("rbind", byLocation)
	return(newDataSS)
}

offline = offline[ offline$mac != "00:0f:a3:39:dd:cd", ]

keepVars = c("posXY", "posX","posY", "orientation", "angle")

onlineCVSummary = reshapeSS(offline, keepVars = keepVars, 
	                          sampleAngle = TRUE)

onlineFold = subset(onlineCVSummary, 
	                  posXY %in% permuteLocs[ , 1])

offlineFold = subset(offlineSummary,
	                   posXY %in% permuteLocs[ , -1])

estFold = predXY(newSignals = onlineFold[ , 6:11], 
	               newAngles = onlineFold[ , 4], 
	               offlineFold, numAngles = 3, k = 3)

actualFold = onlineFold[ , c("posX", "posY")]
calcError(estFold, actualFold)

K = 20
err = rep(0, K)

for (j in 1:v) {
	onlineFold = subset(onlineCVSummary, 
	                    posXY %in% permuteLocs[ , j])
	offlineFold = subset(offlineSummary,
	                     posXY %in% permuteLocs[ , -j])
	actualFold = onlineFold[ , c("posX", "posY")]
	
	for (k in 1:K) {
	  estFold = predXY(newSignals = onlineFold[ , 6:11],
	                   newAngles = onlineFold[ , 4], 
	                   offlineFold, numAngles = 3, k = k)
	  err[k] = err[k] + calcError(estFold, actualFold)
	}
}

pdf(file = "Geo_CVChoiceOfK.pdf", width = 10, height = 6)
oldPar = par(mar = c(4, 3, 1, 1))
plot(y = err, x = (1:K),  type = "l", lwd= 2,
	   ylim = c(1200, 2100),
	   xlab = "Number of Neighbors",
	   ylab = "Sum of Square Errors")

rmseMin = min(err)
kMin = which(err == rmseMin)[1]
segments(x0 = 0, x1 = kMin, y0 = rmseMin, col = gray(0.4), 
	       lty = 2, lwd = 2)
segments(x0 = kMin, x1 = kMin, y0 = 1100,  y1 = rmseMin, 
	       col = grey(0.4), lty = 2, lwd = 2)

#mtext(kMin, side = 1, line = 1, at = kMin, col = grey(0.4))
text(x = kMin - 2, y = rmseMin + 40, 
	   label = as.character(round(rmseMin)), col = grey(0.4))
par(oldPar)
dev.off()

estXYk5 = predXY(newSignals = onlineSummary[ , 6:11], 
	               newAngles = onlineSummary[ , 4], 
	               offlineSummary, numAngles = 3, k = 5)

calcError(estXYk5, actualXY)

predXY = function(newSignals, newAngles, trainData, 
	                numAngles = 1, k = 3){
	
	closeXY = list(length = nrow(newSignals))
	
	for (i in 1:nrow(newSignals)) {
	  trainSS = selectTrain(newAngles[i], trainData, m = numAngles)
	  closeXY[[i]] = findNN(newSignal = as.numeric(newSignals[i, ]),
	                         trainSS)
	}

	estXY = lapply(closeXY, function(x)
	                          sapply(x[ , 2:3], 
	                                  function(x) mean(x[1:k])))
	estXY = do.call("rbind", estXY)
	return(estXY)
}


```

##Appendix - Code  

## References  

